Factoring with $n+2$ clean qubits and $n-1$ dirty qubits

### 2017-05-28 doc/paper.tex
Table

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\title{Factoring with $n+2$ clean qubits and $n-1$ dirty qubits}
\author{Craig Gidney}

@@

In figure \ref{fig:dependencies} we show a dependency graph of the constructions discussed in this paper.
The total cost of the period-finding step in Shor's algorithm, using our full construction, is $O(n^3 \lg n)$ gates and $O(n^3)$ depth.
We perform $O(n)$ modular multiplications, each of which uses $O(n)$ modular additions and offsets, each of which uses $O(n \lg n)$ gates \cite{haner2016} and $O(n)$ depth.

Our main improvements over previous constructions are 1) the use of pivot-flips for modular addition, 2) the use of bimultiplication for modular multiplication, and 3) the $O(n)$ incrementer requiring only a single dirty ancilla.

Previous modular addition constructions worked by temporarily storing an is-wraparound-needed comparison in a clean ancilla \cite{takahashi2006, haner2016}.
Pivot flips also require ancillae (two, actually), but the ancillae can be dirty and, in the context of period-finding, there are always qubits available to borrow whenever a pivot-flip is needed.
This improvement saves a qubit in the overall circuit, reducing the total number of qubits required for period finding (on modular exponentiation against an $n$-bit modulus) from $2n+2$ to $2n+1$.

Previous modular multiplication constructions required a clean ancilla register \cite{haner2016}.
We use an an extra multiply-accumulate and negation to prevent the ancilla register from leaking into the work register, and add a clean-up multiplication informed by the work register at the end of the circuit to undo any effects on the ancilla register.
This improvement allows $n-1$ of the qubits in the ancilla register to be dirty (only the MSB must be zero; to ensure the register's value is in range).

Previous published incrementers (not counting an unpublished version of our construction \cite{gidney2015} cited by \cite{haner2016}) required either $O(n^2)$ gates or $\omega(1)$ ancillae \cite{draper2000, barenco1995}.
Our classical incrementer constructions use $O(n)$ gates and a single dirty ancilla.

Note that, for classical reversible computation, 1 dirty ancilla is the minimum possible for an incrementer.
An increment operation on $n$ bits is equivalent to performing $2^n-1$ state-swaps that move the state $|0\rangle$ from the bottom of the state-space to the top one step at a time.
Note that $2^n-1$ is odd for non-trivial $n$, and therefore the parity of the state permutation performed by an increment operation is odd.
However, the parity of the permutation performed by any classical gate that doesn't cover the entire circuit is even.
Since odd-parity permutations can't be implemented by composing even-parity permutations, it is impossible to perform an increment operation (that involves every bit) with constant-sized operations.
An unused bit (i.e. a dirty ancilla) must be present.

When quantum operations are available, an increment can be performed without any ancilla.
The asymptotic cost is the same, but the constant factors are worse.
See figure \ref{fig:bootstrap-ancilla} for the construction.

>>>>

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{pbox}
\title{Factoring with $n+2$ clean qubits and $n-1$ dirty qubits}
\author{Craig Gidney}

@@

\begin{figure*}
  \centering
  \makebox[\linewidth]{
    \begin{tabular}{ l | c c c c c l }
       & Year & Depth & Gates & Clean Qubits & Total Qubits \\
      \hline
      Shor \cite{Shor1999} & 1994 & $\Theta(n M(n))$ & $\Theta(n M(n))$ & $\Theta(n)$ & $\Theta(n)$ \\
      Beckman et al. \cite{beckman1996} & 1996 & $\Theta(n^3)$ & $\Theta(n^3)$ & $5n + 1$ & $5n + 1$ \\
      Veldral et al. \cite{vedral1996} & 1996 & $\Theta(n^3)$ & $\Theta(n^3)$ & $4n + 3$ & $4n + 3$ \\
      Beauregard \cite{beauregard2003} & 2003 & $\Theta(n^3)$ & $\Theta(n^3 \lg n)$ & $2n+3$ & $2n+3$ \\
      Takahashi et al. \cite{takahashi2006} & 2006 & $\Theta(n^3)$ & $\Theta(n^3 \lg n)$ & $2n+2$ & $2n+2$ \\
      (ours) & 2017 & $\Theta(n^3)$ & $\Theta(n^3 \lg n)$ & $n+2$ & $2n+1$ \\
    \end{tabular}
  }
  \caption{\emph{•}
    Space-efficient constructions of Shor's algorithm over time.
    The table only includes papers that improved the number of qubits required to perform the algorithm and gave explicit circuit constructions.
    $M(n)$ is the classical time-complexity of multiplication, which is known to be asymptotically at most $n \cdot (\lg n) \cdot 2^{O(\lg^* n)}$ \cite{furer2007}.
  }
  \label{fig:table-over-time}
\end{figure*}

Recall that figure \ref{fig:dependencies} shows a dependency graph of the constructions discussed in this paper.
Our asymptotic costs are dominated by performing $O(n)$ modular multiplications, each of which uses $O(n)$ modular additions and offsets, each of which uses $O(n \lg n)$ gates \cite{haner2016} and $O(n)$ depth.
The total cost of the period-finding step in Shor's algorithm, using our construction, is $O(n^3 \lg n)$ gates and $O(n^3)$ depth.

In figure \ref{fig:table-over-time} we show how the number of qubits needed for Shor's algorithm has improved over time.
The previous improvement, from $2n+3$ to $2n+2$, occurred over a decade ago.

Our main improvements over previous arithmetic constructions are 1) the use of pivot-flips for modular addition, 2) the use of bimultiplication for modular multiplication, and 3) the $O(n)$ incrementer requiring only a single dirty ancilla.

Previous modular addition constructions worked by temporarily storing an is-wraparound-needed comparison in a clean ancilla \cite{takahashi2006, haner2016}.
Pivot flips also require ancillae, but the ancillae can be dirty and, in the context of Shor's algorithm, there are always qubits available to borrow whenever a pivot-flip is needed.
This improvement ends up saving a qubit, reducing the total number of qubits required to perform Shor's algorithm from $2n+2$ to $2n+1$.

Previous modular multiplication constructions required a clean ancilla register \cite{haner2016}.
We use an an extra multiply-accumulate and negation to prevent the ancilla register from leaking into the work register, and add a clean-up multiplication informed by the work register at the end of the circuit to undo any effects on the ancilla register.
This improvement allows $n-1$ of the qubits in the ancilla register to be dirty.
Only the MSB has to stay clean, to ensure the register's value is less than the modulus being factored.

Previous published incrementers (not counting an unpublished version of our construction \cite{gidney2015} cited by \cite{haner2016}) required either $O(n^2)$ gates or $\omega(1)$ ancillae \cite{draper2000, barenco1995}.
Our classical incrementer construction uses $O(n)$ gates and a single dirty ancilla.

Note that, for classical reversible computation, 1 dirty ancilla is the minimum possible for an incrementer.
An increment operation on $n$ bits is equivalent to performing a state-permutation that uses $2^n-1$ swaps to move the state $|0\rangle$ from the bottom of the state-space to the top of the state-space one step at a time.
Note that $2^n-1$ is odd for non-trivial $n$, and therefore the parity of the state permutation performed by an increment operation is odd.
However, the parity of the permutation performed by any classical gate that doesn't cover the entire circuit is even.
Since odd-parity permutations can't be implemented by composing even-parity permutations, it is impossible to perform an increment operation that covers every bit with constant-sized operations.
An uncovered bit (i.e. a dirty ancilla) must be present.

When quantum operations are available, the parity barrier can be bypassed and incrementing can be performed without any ancilla.
The asymptotic cost is the same, but the constant factors are worse.
See figure \ref{fig:bootstrap-ancilla} for the construction.
